---
title: "Missing values"
author: "Juan Rocha"
date: "May 2021"
output:
  html_notebook:
      toc: true
      toc_float: true
      highlight: tango
      code_folding: hide
      theme: 
          code_font:
            google: Fira Code
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(naniar)
library(plotly)
library(patchwork)

theme_set(theme_light())
```

## Ecological footprint

```{r}
ecofp <- list()
fls <- list.files("data/ecofootprint/")
ecofp <- map(fls, function(x) {
    jsonlite::read_json(
        path = paste0("data/ecofootprint/", x),
        simplifyVector = TRUE)
    }) %>% 
    bind_rows()

ecofp
```



```{r}
ecofp %>% 
    ggplot(aes(year, value)) +
    geom_line(aes(color = shortName), show.legend = FALSE) +
    facet_wrap(~record, scales = "free_y" ) +
    theme_light(base_size = 9)
```

```{r}
ecofp %>%
    filter(record == "EFConsPerCap") %>%
    mutate(missing = is.na(value)) %>% 
    ggplot(aes(year, countryName)) +
    geom_tile(aes(fill = value)) +
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.y = element_text(size = 4), legend.position = "top")
```

```{r}
g1 <- ecofp %>%
    filter(record == "EFConsPerCap") %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(year) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    ggplot(aes(year, n_nas)) + 
    geom_col(aes(fill = (year>1995)), alpha = 0.5) + 
    coord_flip() +
    theme_light(base_size = 7) +
    theme(legend.position = c(0.8,0.8)) + labs(title = "Missing values per year")

g2 <- ecofp %>%
    filter(record == "EFConsPerCap", year > 1995) %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(countryName) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    filter(n_nas > 0) %>% 
    mutate(countryName = as_factor(countryName) %>% fct_reorder(., n_nas)) %>% 
    ggplot(aes(countryName, n_nas)) + 
    geom_col(aes(fill = (n_nas < 11)), alpha = 0.5) + coord_flip() +
    theme_light(base_size = 7) + 
    labs(title = "Missing values per country after 1995") +
    theme(legend.position = c(0.8, 0.1))

g1 + g2
```

An alternative is use the data after 1995 where `NA`s start decreasing but still give us a couple of decades worth of data to work with. Another is limiting to countries where the count of `NA`s is less than 10.

## Gini: Inequality

```{r}
inq <- readxl::read_xlsx(
    path = "~/Documents/Projects/DATA/WIID_World_Income_Inequality_DB/WIID_06MAY2020.xlsx")


inq %>% 
    #filter(resource == "Income (net/gross)") %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    facet_grid(region_un ~ resource_detailed ) +
    theme_light(base_size = 6)

```
Need to read the docs carefully, it seems many measurements per year and it's hard to tell what is the unit per time that we should use for the analysis.

```{r}
inq %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    facet_grid(region_un ~ resource)
```
It seems that `Income (net)` is the variable with most observations. Let's see:

```{r}
g <- inq %>% 
    filter(year > 1950, resource == "Income (gross)") %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), size = 0.1, show.legend = FALSE) +
    facet_wrap(~resource_detailed)

g
```
```{r}
inq %>% 
    filter(year > 1950, resource_detailed == "Income, gross") %>% 
    ggplot(aes(gini_reported, palma)) +
    geom_miss_point(size = 1, alpha = 0.5) +
    theme(legend.position = c(0.1,0.8))
```

```{r}
inq %>% 
    filter(year > 1950, resource_detailed == "Income, gross") %>% 
    filter(country == "Australia")
```
Note there are multiple estimates per year-country because the database report different sources. 

```{r}
inq %>% 
    filter(year > 1950, resource_detailed == "Income, gross") %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini") %>% 
    mutate(year = as.numeric(year)) %>% 
    ggplot(aes(year, country)) +
    geom_tile(aes(fill = gini)) +
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.y = element_text(size = 4), legend.position = "top")
```
```{r}
inq %>% 
    filter(!is.na(palma), year > 1990) %>% select(country, year, palma) %>% 
    group_by(country, year) %>% 
    summarize(palma_mean = mean(palma, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = palma_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "palma_mean") %>%
    mutate(year = as.numeric(year)) %>%
    ggplot(aes(year, country)) +
    geom_tile(aes(fill = palma_mean)) +
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.y = element_text(size = 4), legend.position = "top")
```
To be fair with Gini, let's to the same over all Gini's reported. Even if Gini comes computed by different studies and in different currencies, the coefficient is always between 0-100, meaning it is comparable.

```{r}
inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year)) %>%
    ggplot(aes(year, country)) +
    geom_tile(aes(fill = gini_mean)) +
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.y = element_text(size = 4), legend.position = "top")
```
So how many country / years can we use? Remember the data has `r inq$country %>% unique() %>% length()` countries

```{r}
inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    group_by(year) %>% 
    summarise(n_nas = sum(missing), missingness = n_nas / 200)
```
We are dealing with ~50% of missing values on the best years, at best we can work with 100 countries. Which ones?

```{r}
inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    group_by(country) %>% 
    summarise(n_nas = sum(missing), missingness = n_nas / (2018-1990)) %>% 
    arrange(missingness)
```

If we allow for 60% missing values, we get 80 countries, 30% NAs is ~60 countries. If needed one can impute `NA`s with a mean trend. But how much is that messing with the real trends? Double click on the legend to select particular countries:

```{r fig.width=7}
g <- inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    ggplot(aes(year, gini_mean)) +
    geom_line(aes(color = country), show.legend = FALSE, size = 0.5) 
    # geom_miss_point(show.legend = FALSE)
ggplotly(g)
```

This justifies why the step-wise approach that Tong first proposed makes sense. If we restrict the analysis to good-data countries, we miss key actors like South Africa, the country with highest inequality in the world. 

## GNI: Prosperity

```{r}
dat1 <- read_csv(
    "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGData.csv") %>% 
    janitor::clean_names()
```

```{r}
dat1 %>% 
    select(-x33) %>% 
    filter(str_detect(indicator_name, "GNI")) %>% 
    pivot_longer(cols = x1990:x2017, names_to = "year", values_to = "value") %>% 
    mutate(year = str_remove(year, "x")) %>% 
    mutate(year = as.numeric(year)) %>% 
    ggplot(aes(year, value)) +
    geom_line(aes(color = country_name), show.legend = FALSE) +
    #geom_miss_point(show.legend = FALSE) +
    facet_wrap(~indicator_name, scales = "free_y")
```

```{r}
## How many missing values? 10506, but who has more?
missingness <- dat1 %>% 
    select(-x33) %>% 
    filter(str_detect(indicator_name, "GNI")) %>% 
    pivot_longer(cols = x1990:x2017, names_to = "year", values_to = "value") %>% 
    mutate(year = str_remove(year, "x")) %>% 
    mutate(year = as.numeric(year)) %>% 
    mutate(ismissing = is.na(value)) %>% 
    group_by(country_code, indicator_name, country_name) %>% 
    summarize(missingness = sum(ismissing)/n())

missingness %>% 
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missingness)) + 
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.x = element_text(angle =90), legend.position = "top")
```

Decide who to exclude and which variable to use

```{r}
missingness %>% 
    ungroup() %>% 
    group_by(indicator_name) %>% 
    summarize(x = mean(missingness)) %>% 
    arrange(x)
```

```{r}
missingness %>% 
    ungroup() %>% 
    group_by(country_name) %>% 
    summarize(x = mean(missingness)) %>% 
    arrange(desc(x)) 
```
It seems that `GNI per capita (current US$)` is the one with less missing values. Tong recomments using the one with Atlas method wich is close on missingness, only 18%. A threshold of 50% missigness per country will left out places like Somalia, Iraq, Costa Rica, Zambia, China or Chile. Perhaps 60% is tolerable? 
