---
title: "Subnational inequality"
author: "Juan Rocha"
date: "Updated `r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE, warning = FALSE, message = FALSE, echo = TRUE,
  fig.width = 8, fig.height = 5
)


library(tidyverse)
#library(naniar)
#library(plotly)
library(patchwork)
library(fs)
library(tictoc)

# time series
# library(tsibble)
# library(fable)
# library(imputeTS)
# library(slider)

## spatial
library(terra)
library(tidyterra)
library(ggspatial)
library(stars)
library(sf)

# # clustering
# library(vegan)
# library(NbClust)
# library(clValid)
# library(mclust)

#theme_set(theme_light(base_size = 12))
```

## Inequality

Exploring subnational inequality data from Kummu et al. Load data:

```{r}

fls <- dir_ls("~/Documents/Projects/DATA/Kummu_inequality/transfer_294976_files_744c217c/")
dat <- rast(fls[2]) # as raster
dat$gini_disp_1990 |> plot()

dat <- vect(fls[5]) # as vector from terra
#terra::plot(dat$gini_disp_2021)

#dat <- sf::read_sf(fls[5])

tmap::tm_polygons(dat)

ggplot(data = dat) +
    geom_spatvector(aes(fill = `2020`), color = "white", linewidth = 0.1) +
    scale_fill_viridis_c()
   # scale_fill_gradient2() 
   # 
dat |> as_tibble() |> 
    ggplot(aes(`1990`, `2021`)) +
    geom_point(aes(color = slope)) +
    geom_abline(slope = 1, intercept = 0) +
    scale_color_gradient2()
```

Time series

```{r}
dat |> 
    as_tibble() |> 
    pivot_longer(cols = `1990`:`2021`, names_to = "year", values_to = "Gini") |> 
    ggplot(aes(year, Gini)) +
    geom_line(aes(group = iso3), alpha = 0.5)
```

## Human footprint

```{r}
hfp <- dir_ls("/Users/juanrocha/Documents/Projects/DATA/HII_human_impact_index/first-gen/")

fp1 <- stars::read_stars(hfp[10])
fp1 <- rast(hfp[20])

tic()
terra::plot(fp1, type = "continuous", #col = map.pal("viridis", 100),
            #range = c(-100,5000),
            col = c("#224f1a","#a3ff76","#feff6f","#a09568","#ffa802","#f7797c","#fb0102","#d87136","#a90086","#7a1ca5","#421137","#000000")
            )
toc() #18s
#click(n=1)

# tic()
# m <- max(fp1, na.rm = TRUE)
# toc() # 496s

m

fp1[,10000,] |> range(na.rm= TRUE)
```

From Google Earth engine 
```{javascript eval=FALSE}
var hiiviz = {min: 5, max: 5000, palette: ["224f1a","a3ff76","feff6f","a09568","ffa802","f7797c","fb0102","d87136","a90086","7a1ca5","421137","000000"]
```

### Spatial joints

```{r eval = FALSE, include = FALSE}
# library(furrr)
# library(listenv)
# plan(multisession, workers = 10)
## this fails
out <- list()
out <- map(
    .x = hfp,
    .f = function(x){
        tic("extracting values")
        dat <- extract(x = rast(eval(x)), y = dat, fun = mean, bind = TRUE, na.rm = TRUE)
        toc() # 7121.054, 2hrs with terra
        return(dat)
    },
    .progress = TRUE
) # >20hrs

df_dat <- out |> 
    map(.f = function(x) st_as_sf(x) |> select(-c(admID:p.value))) 

save(df_dat, file = "data/subnational_hii.Rda")

df_dat <- df_dat |> 
    map(function(x) as.data.frame(x) |> select(starts_with("hii_"))) |> 
    bind_cols() |> 
    as_tibble()

bind_cols(dat, df_dat)


# out <- listenv()
# tic()
# for (i in seq_along(hfp)){
#     hfp; dat;
#     tic("extracting values")
#     out[[i]] %<-% terra::extract(
#         x = rast(hfp[i]),
#         y = dat,
#         fun = mean, na.rm = TRUE,
#         bind = TRUE
#     )
#     toc()
# }
# toc()
# 
# out[[2]]

# plan(sequential)
# tic("extracting values")
# dat <- extract(x = rast(hfp[1]), y = dat, fun = mean, bind = TRUE, na.rm = TRUE)
# toc() # 7121.054, 2hrs with terra

## playing with stars:: functions
# dat needs to be read with sf
# tic()
# fp1 <- aggregate(x = fp1, by = dat, FUN = mean, na.rm = TRUE)
# toc()
```

In the future, no need to run previous chunk

```{r}
load("data/subnational_hii.Rda")
```



Mix all but keep it as a data frame:
```{r}
dat <- dat |> as.data.frame() |> select(-slope, -estimate, -p.value) |> 
    as_tibble() 

df_dat <- df_dat |> 
    add_column(admID = dat$admID, iso3 = dat$iso3)

dat <- dat |> pivot_longer(cols = `1990`:`2021`, names_to = "year", values_to = "gini") 
df_dat <- df_dat |> 
    pivot_longer(cols = starts_with("hii"), names_to = "year", values_to = "hii")

df_dat <- df_dat |> 
    mutate(year = str_remove(year, "hii_")) |> 
    mutate(year = str_remove(year, "\\.01\\.01$")) |> 
    mutate(year = as.numeric(year))

dat <- mutate(dat, year = as.numeric(year))

tlm <- left_join(
    df_dat |> filter(!is.na(hii)), 
    dat |> filter(!is.na(gini)),
    by = c("admID", "iso3", "year"))
```

add the proxy for prosperity
```{r}
prosp <- vect(fls[6])
prosp <- prosp |> 
    as.data.frame() |> 
    as_tibble() |> 
    select(-slope,-estimate, -p.value) |> 
    pivot_longer(cols = `1990`:`2021`, names_to = "year", values_to = "gnic") |> 
    mutate(year = as.numeric(year))

prosp |> select(admID, iso3) |> 
    unique()

tlm |> select(admID, iso3) |> 
    unique()

tlm <- tlm |> left_join(prosp)
### Not the same geo codes, ask Matti
```

```{r}
tlm |> 
    ggplot(aes(hii, gini)) +
    # geom_path(aes(group = iso3), alpha = 0.2)
    geom_density2d_filled()

tlm |> ggplot(aes(gini)) + geom_density()

tlm |> 
    group_by(iso3) |> 
    summarize(gini = mean(gini, na.rm = TRUE)) |> 
    ggplot(aes(gini)) + geom_density()


```

Check differences in spatial units:
```{r}
gdi <- vect(fls[4])
prosp <- vect(fls[6])
ineq <- vect(fls[5])
gdi
ggplot(data = ineq) +
    geom_spatvector(aes(fill = `2020`), color = "white", linewidth = 0.1) +
    scale_fill_viridis_c()

gdi # 1816 polygons
prosp # 1816 polygons
ineq # 1929 polygons

gdi_units <- gdi |> as.data.frame() |> as_tibble() |> 
    select(admID, iso3) 
ineq_units <- ineq |> as.data.frame() |> as_tibble() |> 
    select(admID, iso3) 


gdi$iso3[!gdi$iso3 %in% ineq$iso3]
ineq$iso3[!ineq$iso3 %in% gdi$iso3]

full_join(
    gdi_units|> group_by(iso3) |> summarize(n_gdi=n()),
    ineq_units|> group_by(iso3) |> summarize(n_inq=n())) |> 
    mutate(both = n_gdi == n_inq) |> 
    print(n=252)
```

I'll need to use the raster layer and compute the vector one using the same layer.

```{r}
rst <- rast(fls[2])
tic()
ineq2 <- extract(rst, gdi, fun = mean, ID = TRUE)
toc() # 47s

ineq2 <- ineq2 |> as_tibble() |> 
    add_column(admID = gdi$admID, iso3 = gdi$iso3) 

ineq2 # 1816 rows

ineq2 |> 
    pivot_longer(cols = starts_with("gini"), names_to = "year", values_to = "gini") |> 
    mutate(year = str_remove(year, "gini_disp_"),
           year = as.numeric(year)) |> 
    select(-ID)

## Here I lose 20k obs, ineq2 has 

tlm <- tlm |> left_join(
    ineq2 |> 
    pivot_longer(cols = starts_with("gini"), names_to = "year", values_to = "gini") |> 
    mutate(year = str_remove(year, "gini_disp_"),
           year = as.numeric(year)) |> 
    select(-ID)
)
    
```

