---
title: "Biosphere Stewardship in an Unequal World: A Sustainable Development Goals Approach"
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: refs.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)

## word counts
## Abstract 150
## Intro 697
## Method 957
## Results 621
## Discussion 1434
## Conclusion 267
## Total main 3976 + 150 abstract
## 
## Targets: 
# 
# - NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
# 70max
# 
# - WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
# abstract

# Notes:

# how to make it different from the trilemma?# 
# clearer story
# how much of the patterns hold constant if we remove indicators?
# sensitivity analysis for MFA / ordination and clustering.
```

# Outline

Target: EcolEcon?
/ People and Nature?
/ PNAS / ERL / NatSust?
[4000w]

Authors:

Introduction {800w}

-   Previous studies proposed the hypothesis of the trilemma

-   Why inequality and the biosphere / as opposed to other SDGs

-   Do the trade-offs hold at national scale with SDG data?

Methods {800w}:

-   SDG datasets at national scale collected by World Bank and United Nations
-   Selection of variables
-   Multiple Factor Analysis (PCA over time)
-   Clustering
-   Trajectories

Results {1000w}:

-   We confirm the results of Wu with three clusters of countries in two different SDGs datastets.
-   We confirm the existence of inequality traps or regime shifts in inequality.
-   We confirm the path dependency: the pathway from one regime to another is not random and offer insights on the type of policies to favour to achieve multiple goals. It also points to risks that need to be manage to avoid trade-offs.

Discussion {1000w}

-   Poor data quality overall.
    Biodiversity metrics either don't change or are focused on conservation.
    Thus it speaks of intentions not actual state variables.

-   Despite data challenges we confirm the existence of inequality regimes.

-   Some of the mechanisms by which inequality and the biosphere might be coupled occur at the scale of individuals or households.

-   Future research gaps: do the patterns hold at subnational scales?
    Do the patterns hold with biodiversity variables that actually reflects the state of biodiversity?
    Would the patterns be the same at local scales?
    Would they be different in human dominated systems such as cities?

Conclusion {400w}

SM

-   Figure with missing values from WB dataset: **improve resolution**

-   Figure with missing values from UN dataset: **improve resolution**

-   Correlograms

\pagebreak

\begin{abstract}

  

\end{abstract}

# Introduction

Countries around the world have committed to achieve 17 sustainable development goals as proposed by the United Nations.
The ambitious agenda is materialized in 169 targets and indicators, yet not all targets are monitored or properly measured, and not all countries report them, presenting significant data and knowledge gaps [@min2024].
An open question in sustainability science is if these targets are achievable simultaneously?
Previous work on synergies and trade-offs between sustainable development goals (SDGs) suggest that both options are possible [@schmidt-traub2017; @sachs2024; @pradhan2017; @kroll2019], yet this work is limited to expert elicitation and correlational studies.
It is unclear if there are system or structural constraints that limit a country's capacity to achieve particular SDGs, or if there are milestones that need to be unlocked before it can achieve the desired performance in a particular dimension of sustainability.
In other words, these correlations do not account for the path dependency and causal inference inherent on the process of development.

When the SDGs were adopted in 2015, the United Nations began to create a comprehensive database of SDG indicators [@schmidt-traub2017; @sachs2024], to be updated annually.
Based on these data, previous analyses found that goals associated with poverty alleviation, well-being, economic development, and innovation (e.g. SDGs 1, 3, 7, 8, 9) tended to be synergistic with other goals, while goals related to responsible consumption, climate action, natural resources, and cooperation showed the most trade-offs (e.g. SDGs 11, 12, 13, 14, 16, 17) [@pradhan2017; @kroll2019].
Drawing upon relevant World Bank data, Lusseau and Mancini [@lusseau2019] find these patterns to be modulated by the overall income level of countries, with low-income countries showing synergies across all goals, while trade-offs start to appear in higher income categories.
More recently, [@xiao2024] analysed transboundary SDG interactions and found that high income countries play a disproportionate role in influencing the achievement of SDGs in other countries.
These differences suggest that it may be worthwhile investigating if the nature of the mechanisms linking SDG to each other could be dependent on income level.

Here we address the question of synergies and trade-offs by focusing on goals related to biosphere and inequality (SDGs 5, 6, 10, 13, 14 and 15).
The motivation is threefold.
First, the Convention for Biological Diversity is currently negotiating and agreeing on the next set of goals and ambitions to bend the curve of biodiversity loss.
Understanding which countries are well positioned to achieve biodiversity goals or not will be necessary to design strategies that support countries with low capabilities, that address the dimensions of justice, and design targets with plausibility in mind [@díaz2020].
Second, within-country inequality has been rising in the last decades even in high income countries [@milanovic2024], and as a consequence of the Covid pandemic, in between country inequality raised for first time in a generation [@wade2023world].
These statistics set back progress on the inequality SDG by at least a decade, and undermines the mantra of leaving no one behind.
And third, recent conceptual and theoretical work has proposed mechanisms by which increase in inequality can impact the environment, while changes in the environment can feedback into further inequalities [@hamann2018; @dade2022; @leach2018].

Recent work also shows that a trilemma exists where countries struggle to simultaneously achieve high prosperity, high environmental standards while reducing inequality [@Wu_2024].
Using data from environmental footprint, the gross national product, and the Gini coefficient time series, 53 countries were clustered and a typology of trajectories identified.
No country achieved the three goals simultaneously, and Latin American countries seem to have dynamics characterized by an inequality trap.
Some countries trajectories on the trilemma space, however, suggest the hypothesis that social progress can be achieved without compromising the biosphere [@Wu_2024].
While no country has simultaneously achieved these three goals, some countrie are indeed moving on the right direction [@Wu_2024; @chrisendo2024; @liu2025]

It remains open questions whether these patterns are robust across different datasets and scales, or whether there are specific driving factors and feedback mechanisms that underlie inequality traps.
The dichotomy between low-income countries exhibiting synergies across all goals, while trade-offs starting to appear in higher-income countries [@lusseau2019] motivates the need to study mechanisms explaining these trade-offs and how they may differ due to countries’ level of income.
If the hypothesis of the inequality trap is true, we should observe bimodal or multimodal distributions in inequality and environmental variables.
Each mode would correspond to an inequality regime, and the transitions probability of staying within one regime should be much higher than the probability of shifting regimes.
If there are nonlinear dynamics in inequality keeping countries trapped on a particular regime, then we should also observe hysteresis or different break points between regimes.
Here we address such questions by studying synergies and trade-offs between SDGs related to inequality and the biosphere.

# Methods

```{r echo = FALSE, message=FALSE, warning=FALSE}
# dat <- read_csv(file = "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGData.csv") |> 
#     janitor::clean_names()
# 
# un_countries <- read_csv2(file = "~/Documents/Projects/DATA/SDGs_UNStats/UNSD — Methodology.csv")

# old UN dataset
# sdg <- read_csv(file = "~/Documents/Projects/DATA/SDGs_UNStats/20180817102044132_juan.rocha@su.se_data.csv") |> janitor::clean_names()

## new UN dataset:
# fls <- fs::dir_ls(paste0(here(), "/data/sdgs/"))
# 
# sdg <- list()
# 
# for (i in seq_along(fls)){
#     load(fls[i])
#     sdg[[i]] <- d
# }
# 
# 
# sdg <- sdg |> bind_rows() |> 
#     unnest(c(goal, target, indicator))
# 
# sdg <- sdg |> 
#     left_join(
#         select(un_countries, country = `Country or Area`, starts_with("ISO")),
#         by = c("geo_area_name" = "country")) |> 
#     filter(value != "-")|> 
#     mutate(value = as.numeric(value))

```

*Datasets:* To that end, we used the SDGs datasets made openly available by the World Bank (WB) and United Nations (UN).
The WB dataset offers 403 indicators, with time series from 1990 to 2019 for 263 countries or administrative areas (N = 2'013.791 observations).
The UN dataset offers time series from 1963 to 2025 for 17 SDGs, 168 targets, and 247 indicators, 687 time series, in 413 administrative areas (N = 2'821.669 observations).
Despite their coverage, both datasets contain a high proportion of missing values and the sample is unbalanced across countries and time, meaning some countries have better temporal coverage than others.
We focused subsequent analysis only on country level indicators for SDGs 5, 6, 10, 13, 14 and 15, related to inequality and the biosphere respectively.
We complemented the SDGs datasets with inequality data from the World Inequality database (WID), using their estimates of the ratio of pre-tax national income for working adults (population \> 20 years old) computed as the share of the top 10% over the share of the bottom 50% (`rptinc992j_p0p100`), the share of the 1% (`sptinc992j_p99p100`), and the Gini coefficient (`gptinc992j_p0p100`).
We also used their estimates for net wealth inequality computed as the share of the top 10% over the share of the bottom 50% (`rhweal992j_p0p100`), the share of the 1% (`shweal992j_p99p100`), and the Gini coefficient(`ghweal992j_p0p100`).

```{r echo = FALSE}
#rm(dat, sdg, fls) # remove unnecessary objects
load(file = paste0(here(), "/data/cleaned_SDGs_2025.Rda")) 
un_dat <- ungroup(un_dat)
wb_dat <- ungroup(wb_dat)

```

*Variable selection:* We computed the proportion of missing values for all time series related to our initial selection of indicators (Figs S1, S2, S3).
We discarded indicators for which time series contained more than 30% of missing values, or less than 45 countries.
Missing values were then imputed using a cubic spline, leaving us with `r wb_dat |> pull(country_code) |> unique() |> length()` countries, `r wb_dat |> pull(year) |> unique() |> length()` years of data across 9 indicators for the WB dataset; and `r un_keys |> filter(series %in% unique(un_dat$series)) |>  pull(series) |> unique() |> length()` series capturing `r un_keys |> filter(series %in% unique(un_dat$series)) |>  pull(indicator) |> unique() |> length()` indicators, `r un_dat |> pull(geoAreaName) |> unique() |> length()` countries over `r un_dat |> pull(timePeriodStart) |> unique() |> length()` years for the UN dataset.
The UN dataset was further reduced to 67 countries because the WID does not report inequality time series for Fiji.
Table \ref{tbl1} summarizes our selected variables, their units and available ranges.
A list of the countries analysed is presented in the supplementary information (SI).

*Ordination:* We used multiple factor analysis (MFA) and principal component analysis (PCA) to reduce the dimensionality of the data and explore similarities and differences across countries.
MFA enables us to specify the nested structure of our data and account for repeated observations of our variables over time.
We recover some of the qualitative results with PCA as robustness check, but these results are presented in the SI.
We also performed a clustering sensitivity analysis following the protocols by Charrad [@charrad2014] and Brooks [@brock2008].
We tested over 10 clustering techniques and compared them across \>30 performance metrics to infer from the data what are the optimal numbers of clusters to fit and preferable algorithms.

*Analysis of trajectories:* With the results from the MFA we identified candidate variables where synergies or trade-offs are expected.
A trade-off in the reduced dimensional space occurs when improving on the direction of one indicator (e.g. reducing inequality) implies a decline in the direction of another indicator.
Similarly, a synergy would be when increasing in one indicator coincides with a country towards improving in another indicator.
We extracted trajectories of the countries of analysis for some of these candidate variables.
If the hypothesis of inequality traps or inequality regimes is true, we would expect to find multimodal distributions in inequality indicators with respect to other variables; we expect the modes of these distributions to be correlated to the countries groups identified via clustering analysis; and we expect that most countries stay within their region of the parameter space and perhaps a few of them able to move between regimes.

```{r table1, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, opts.label="tbl1"}

library(kableExtra)

load(paste0(here(),"/data/wb_var_defs.Rda"))
load(paste0(here(),"/data/wii_small-data.RData"))


un_vars <- un_keys |> 
    filter(series %in% unique(un_dat$series)) |> 
    select(Goal = goal, Series = series, Variable = seriesDescription) |> 
    mutate(Source = "UN")

wb_vars <- wb_key |> 
    filter(indicator_name %in% colnames(wb_dat)[4:12]) |> 
    select(Series = series_code, Variable = indicator_name) |> 
    mutate(Goal = c("15", "10", rep("13", 2), "10", rep("15",2),"10","10" ),
           Source = "WB")

iq_vars <- tibble(
    Source = "WII", 
    Goal = "10", 
    Series = ineq_dat |> 
        unite(col = "series", variable, percentile, sep = "_") |>
        # removed due to zero variance
        filter(!series %in% c("sptinc992j_p0p100","shweal992j_p0p100")) |>
        pull(series) |> unique(),
    Variable = c(
        "Ratio of pre-tax national income for working adults",
        "Ratio of wealth for working adults",
        "Share of the top 1% of wealth",
        "Gini coefficient of wealth",
        "Gini coefficeint of pre-tax income for working adults",
        "Share of the 1% of pre-tax income for working adults"
    )
)


tbl <- bind_rows(wb_vars, un_vars, iq_vars) |> 
    select(Source, Goal, Series, Variable) |> 
    mutate(Goal = as.numeric(Goal)) |> 
    arrange(Goal)

tbl <-  tbl |> mutate(
    Units = str_extract(Variable, "\\(.\\)$"),
    Units = str_remove_all(Units, "\\(|\\)")
) |> mutate(
        Variable = str_remove(Variable, " \\(%\\)"),
        Units = str_extract(Variable, "\\(.*\\)$"),
        Variable = str_remove(Variable, " \\(.*\\)$"),
        Units = str_remove_all(Units, "\\(|\\)"))

tbl$Units[3:4] <- "0 to 1"
tbl$Units[5] <- "US$ per cubic meter"
tbl$Units[6] <- "0 to 1"
tbl$Units[18:23] <- "0 to 1"
tbl$Units[29:32] <- "0 to 1"
tbl$Units[c(7,9,13)] <- "sq. km"

knitr::kable(
    tbl, format = "latex", booktabs = TRUE, caption = "Summary of variables used", label = "tbl1") %>%
  kable_styling(latex_options = c("striped", "hold_position"), 
                full_width = TRUE, font_size = 8) |> 
  column_spec(c(1,2), width = "5mm") |> column_spec(3, width = "25mm") |>
  column_spec(4, width = "8cm") 
```

# Results

Despite the differences in coverage with respect to countries, time, and indicators tracked, the ordination in both data sets results in two clusters of countries (Fig \ref{fig:wb}, Fig \ref{fig:un}).
For the WB data set, countries along the first principal component are differentiated by high levels of economic inequality, high energy intensity but low carbon emissions (positive values of Dim 1, eg. green cluster Mexico: MEX or South Africa: ZAF), versus countries with relatively low gender inequality, high agricultural productivity, high carbon emissions, high urbanization and internet access (negative values in Dim 1, yellow cluster).
Forest related variables have the lowest loading on the first two components and do not change much over time, while access to internet or inclusion of women in leadership roles have the highest variability over time.
The first dimension on the ordination is best explained by variability in the inequality variables including women in buisness, access to internet and electricity, while the second dimension is best explained by urbanization and carbon emissions (Fig \ref{fig:SM_WBord}).

\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4in]{fig_mfa_wb}
\caption{\textbf{Multiple factor analysis with World Bank data} The first 10 principal componets explain 94.8\% of the varation, the first two (A) explain 53\%. The 10 first components were used to cluster 151 countries resulting in two clusters (A). The correlation circle across explanatory variables is presented in (B) along their loadings on the first two components of the ordination. Variables in the legend are ordered and coloured according to the SDGs used (e.g. orange for gender equality, blues for water and sanitation, reds for inequality, and greens for life on land).}
\label{fig:wb}
\end{figure*}

The UN dataset offers a similar ordination, where countries with high values along the first axes and higher values along the second axis have the highest inequalities (e.g Fig \ref{fig:un}).
The inequality variables are highly correlated but also explain large amounts of the variance.
Contrary to the WB data, here forest related variables do show variability over time, but variables related to biodiversity loss (Red list index) or some of the area based indicators for water related SDGs do not change much over the time period of the data (2000-2021).
Places with lower economic inequality also tend to have better opportunities for woman to participate in political decision making.
Interestingly, high values on the red list index correlate with lower levels of inequality.
The first dimension on the ordination is best explained by variability in the inequality, while the second dimension is best explained by key biodiversity areas in terrestrial systems, reservoir statistics, as well as women participation in parliament (Fig \ref{fig:SM_UNord}).

\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4.5in]{fig_UN_ordination}
\caption{\textbf{Multiple factor analysis with United Nations data} The first 10 principal components used in the ordination explain 72\% of the variation, the first two (A) explain 29.9\%. The first 10 components were used to cluster 67 countires resulting in two clusters (A). The correlation circle across explanatory variables is presented in (B) along their loadings on teh first two components of the ordination. Variables in the legend are ordered and coloured according to the SDGs used (e.g. reds are inequality, greens are life in land).}
\label{fig:un}
\end{figure*}

We find evidence of multi-modal distributions in inequality and environmental variables.
This finding supports the hypothesis of the existence of inequality regimes both in income and wealth, particularly when inequality is measured as the share of the top 1% (Fig \ref{fig:bim}).
However, the lack of variability in environmental variables and smaller sample size in the UN data set prevent us to statistically derive modes in the distribution or all our variables (Table \ref{tbl1}).
As a result, we only report bimodal distributions for cereal yields, carbon emissions, and energy intensity (WB data, Fig \ref{fig:bim}}), and the red list index (UN data, Fig \ref{fig:SM_UNbim}).
We also confirm that, as expected, these inequality regimes are related with the country typologies identified.

\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4in]{bimodal}
\caption{\textbf{Inequality regimes} We find bimodal distributions for some dimensions of inequality against environmental factors in the WB dataset. Countries trajectories in A using the same cluster groups as in Fig \ref{fig:un}. Bimodal distributions are more common for the share of the top 1\% than other inequality variables (B). The y-axis has been rescaled to the range 0-1 to ease comparison. Supplementary figure \ref{fig:SM_UNbim} shows a similar plot for the UN data, however lack of variability in the environmental data prevents the identification of multiple modes in the distribution.}
\label{fig:bim}
\end{figure*}

# Discussion

Most countries in the world are committed to achieve progress on sustainable development goals by 2030.
Despite several of these aspirational goals wont be meet by 2030; they concentrate political willingness and action that has been translated into some progress in most countries.
One key question is whether there are synergies and trade-offs between these goals.
Here we addressed this question by investigating SDGs related to inequality and the biosphere (SDGs 5, 6, 10, 13, 14, and 15) through ordination methods.

We find evidence of synergies and trade-offs.
For example, almost all inequality metrics are correlated, except the ratio of wealth which is much more noisy and sometimes anti-correlated with inequality metrics on income.
For example, Sweden traditionally has had low income inequality but high wealth inequality, although the former has been increasing as well.
Similarly, gender equity measured as share of women in parliament is generally better in places with lower income inequality but not necessarily low wealth inequality.
Interestingly, countries with high gender equality perform better in area based indicators of ecological and biological conservation.
Higher levels of urbanization are correlated with improvements on access to electricity and internet usage, but also with higher carbon emissions and lower energy intensity.

We had an explicit interest in the interactions between inequality and biosphere related SDGs.
Biodiversity variables are either limited to conservation, or resource production.
Most of the proxies of ecosystem variables are area based (e.g. % forest area), meaning they change very little over time.
The lack of variability in the SDGs datasets and a large proportion of missing values (Figs \ref{fig:SM_WB_env}, \ref{fig:SM_WB_env2}, \ref{fig:SM_UNdata} ) forced us to drop many indicators.
It also questions whether the indicators currently used are good proxies of progress, or lack of it, in the goals.
If a variable does not change over time at a scale at which information can feedback political decisions, then is not very useful as proxy of progress towards a desired direction.
Other biodiversity related values such as cultural values (intrinsic, recreational, spiritual) are not currently captured by SDG indicators.
An interesting avenue for future research is to design comparable observables that can be monitored by third parties (not reported by governments) that are sensitive enough to capture progress or lack of it towards the goals.
Some examples include the essential biodiversity variables initiative advanced by GeoBON, the human footprint index, or recent developments on functional integrity.
All of them offer time series at varied levels of resolution that enable sub national monitoring.

Despite these limitations, we find evidence for bistability in inequality, further supporting the hypothesis of inequality traps [@Wu_2024].
However, data gaps and lack of variability in biodiversity indicators prevent us to further test features of regime shift dynamics such as hysteresis.
To further test the existence of hysteresis we need a better understanding of the potential mechanisms at the country scale that might be generating the alternative regimes, including a controlling parameter at which the shift from one regime to the other should result on a different break point.
Further work can look into this hypothesis by looking into other datasets or longer time series, or by exploring the plausibility of alternative mechanisms through modelling.
Generative models for wealth inequality exist inspired in biological mechanisms of heredity [@arrow2009], but less is known on inequality mechanisms for income or gender dimensions, although some hypothesis have been put forward [@chancel2020unsustainable].

# Conclusion

Notes:

-   Paper figures:

    -   Plot with PCA selection (2-3 components?) and then change over time of the components. Do clustering on trayectories as well? -- and then add a map with the clusters for each dataset.

# References {.unnumbered}

::: {#refs}
:::

\pagebreak

# Supplementary Material {#sec:SM}

\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}

List of countries analyzed in the UN dataset after removing time series with too many missing values or too few countries: `r un_dat |> pull(geoAreaName) |> unique() |> str_flatten_comma(last = " and ")`

List of countries analyzed in the WB dataset after removing time series with too many missing values or too few countries: `r wb_dat |> pull(country_name) |> unique() |> str_flatten_comma(last = " and ")`

\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{WB_environment_SDGs}
\caption{\textbf{World Bank environmental SDGs} }
\label{fig:SM_WB_env}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{WB_environment_SDGs_othervars}
\caption{\textbf{World Bank environmental SDGs additional variables} }
\label{fig:SM_WB_env2}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{WB_selected_vars}
\caption{\textbf{World Bank selected variables} }
\label{fig:SM_WB_select}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{WB_correlogram}
\caption{\textbf{Correlogram of World Bank selected variables} }
\label{fig:SM_WBcorr}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_SDG_10-14-15}
\caption{\textbf{United Nations SDGs 10, 14 and 15 dataset} }
\label{fig:SM_UNdata}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_selected_vars}
\caption{\textbf{UN selected variables} }
\label{fig:SM_UNselect}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_reduced_vars}
\caption{\textbf{UN reduced variables} }
\label{fig:SM_UNred}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_correlogram}
\caption{\textbf{Correlogram of United Nations selected variables} }
\label{fig:SM_UNcorr}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{sm_wb_ordinations}
\caption{\textbf{Variable importance in ordination on WB data} }
\label{fig:SM_WBord}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{sm_un_ordinations}
\caption{\textbf{Variable importance in ordination on UN data} }
\label{fig:SM_UNord}
\end{figure*}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 3.5in]{bimodal_UN}
\caption{\textbf{Bimodality in UN data} The quality of the data does not allow for detection of bimodality in most UN variables. The lack of variation in biosphere related data does not allow the identification of areas in the parameter space where trajectories converge, except for the red list index, where we find multimodal distributions }
\label{fig:SM_UNbim}
\end{figure*}
