---
title: "Trade-offs between biosphere goals and inequality at the national scale"
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: refs.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

```{r data, include=FALSE}
library(tidyverse)
library(here)

## word counts
## Abstract 150
## Intro 697
## Method 957
## Results 621
## Discussion 1434
## Conclusion 267
## Total main 3976 + 150 abstract
## 
## Targets: 
# 
# - NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
# 70max
# 
# - WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
# abstract
```

# Outline

Target: EcolEcon?
/ People and Nature?
/ PNAS / ERL / NatSust?
[4000w]

Authors:

Introduction {800w}

-   Previous studies proposed the hypothesis of the trilemma

-   Do the trade-offs hold at national scale with SDG data?

Methods {800w}:

-   SDG datasets at national scale collected by World Bank and United Nations
-   Selection of variables
-   Multiple Factor Analysis (PCA over time)
-   Clustering

Results {1000w}:

-   download and check world inequality db again

Discussion {1000w}

Conclusion {400w}

SM

-   Figure with missing values from WB dataset: **improve resolution**

-   Figure with missing values from UN dataset: **improve resolution**

-   Correlograms

-   

# Introduction

Countries around the world have committed to achieve 17 sustainable development goals.
The ambitious agenda is materialized in 169 targets and indicators, yet not all targets are monitored and not all countries report them.
An open question in sustainability science is if these targets are achievable simultaneously?
Previous work on synergies and trade-offs between sustainable development goals (SDGs) suggest that both options are possible, yet this work is limited to expert elicitation and correlational studies.
It is unclear if there are structural constraints that limits a country capacity to achieve particular SDGs, or if there are milestones that need to be unlocked before it can achieve the desired performance in a particular dimension of sustainability.
In other words, these correlations do not account for the path dependency inherent of the process of development.

Here we address the question of synergies and trade-offs by focusing on goals related to biosphere and inequality (SDGs 10, 14 and 15).
The motivation is twofold.
First, the Convention for Biological Diversity is currently negotiating and agreeing on the next set of goals and ambitions to bend the curve of biodiversity loss.
Understanding which countries are well positioned to achieve biodiversity goals or not will be necessary to design strategies that support countries with low capabilities, that address the dimensions of justice, and design targets with plausibility in mind.
Second, within country inequality has been rising in the last decades even in high income countries, and as consequence of the Covid pandemic, in between country inequality raised for first time in a generation.
These statistics set back progress on the inequality SDG by at least a decade, and undermines the mantra of leaving no one behind.

Recent work shows that a trilemma exist where countries struggle to achieve high prosperity, high environmental standards while reducing inequality.
Using data from environmental footprint, the gross national product, and the Gini coefficient time series, 53 countries were clustered and a typology of trajectories identified.
No country achieved the three goals simultenously, and Latin American countries seem to have dynamics characterized by an inequality trap.
Countries trajectories on the trilemma space, however, suggest the hypothesis that social progress can be achieved without compromising the biosphere.
It remains an open questions if these patterns hold across different datasets and scales.
Here we address such gap by studying synergies and trade-offs between SDGs related to inequality and the biosphere.

# Methods

```{r echo = FALSE, message=FALSE, warning=FALSE}
dat <- read_csv(file = "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGData.csv") |> 
    janitor::clean_names()

un_countries <- read_csv2(file = "~/Documents/Projects/DATA/SDGs_UNStats/UNSD — Methodology.csv")
sdg <- read_csv(file = "~/Documents/Projects/DATA/SDGs_UNStats/20180817102044132_juan.rocha@su.se_data.csv") |> janitor::clean_names()

sdg <- sdg |> 
    left_join(
        select(un_countries, country = `Country or Area`, starts_with("ISO")),
        by = c("geo_area_name" = "country")) |> 
    filter(value != "-")|> 
    mutate(value = as.numeric(value))


```

*Datasets:* To that end, we used the SDGs datasets made openly available by the World Bank (WB) and United Nations (UN).
The WB dataset offers `r dat$indicator_name |> unique() |> length()` indicators, with time series from 1990 to 2019 for `r dat$country_name |> unique() |> length()` countries or administrative areas (N = 2'013.791 observations).
The UN dataset offers time series from 2000 to 2019 for 17 SDGs, 112 targets, and 356 indicators, in 310 administrative areas (N = 1'072.731 observations).
Despite their coverage, both datasets contain high proportion of missing values and the sample is unbalanced across countries and time, meaning some countries have better temporal coverage than others.
We focused subsequent analysis only on country level indicators for SDGs 10, 14 and 15, related to inequality and the biosphere respectively.
We complemented the SDGs datasets with inequality data from the World Inequality database (WID), using their estimates of the ratio of pre-tax national income for working adults (population > 20 years old) computed as the share of the top 10% over the share of the bottom 50% (`rptinc992j_p0p100`), the share of the 1% (`sptinc992j_p99p100`), and the Gini coefficient (`gptinc992j_p0p100`).

```{r echo = FALSE}
load(file = paste0(here(), "/data/cleaned_sdg.RData"))
# `df_dat` is the data from World Bank, and `df_un_reduced` is the data from UN SDGs
df_dat <- df_dat |> ungroup()
df_un_reduced <- df_un_reduced |> ungroup()
```

*Variable selection:* We computed the proportion of missing values for all time series related to our initial selection of indicators (Figs S1, S2, S3).
We discarded indicators for which time series contained more than 30% of missing values, or less than 45 countries.
Missing values were then imputed using a cubic spline, leaving us with `r df_dat |> pull(country_code) |> unique() |> length()` countries, `r df_dat |> pull(year) |> unique() |> length()` years of data across 9 indicators for the WB dataset; and `r df_un_reduced |> pivot_longer(cols = 3:last_col(), names_to = "var", values_to= "value") |> separate(var, into = c("year", "series_name"), sep = "_") |>  pull(series_name) |> unique() |> length()` indicators, `r df_un_reduced |> pull(iso3) |> unique() |> length() + 1` countries over `r df_un_reduced |> pivot_longer(cols = 3:last_col(), names_to = "var", values_to= "value") |> separate(var, into = c("year", "series_name"), sep = "_") |> pull(year) |> unique() |> length()` years for the UN dataset.
The UN dataset was further reduced to 50 countries because the WID does not report inequality timeseries for Fiji.
Table X summarizes our selected variables, their units and available ranges.
A list of the countries analysed is presented in the supplementary information (SI).

*Ordination:* We used multiple factor analysis (MFA) and principal component analysis (PCA) to reduce the dimensionality of the data and explore similarities and differences across countries.
MFA enables to specify the nested structure of our data and account for repeated observations of our variables over time.
We recover some of the qualitative results with PCA as robustness check, but these results are presented in the SI.
We also performed a clustering routine following the protocols by Charrad [] and Brooks [].
In short, we tested over 10 clustering techniques and compared them across \>30 performance metrics to infer from the data what are the optimal numbers of clusters to fit and preferable algorithms.

*Analysis of trajectories:* With the results from the MFA we extracted trajectories of the countries of analysis on the reduced dimensions space.
We were particularly interested on the principal components related to inequality and performance on SDGs related to life on land and water.
The use of trajectories in the MFA space allows us to test whether countries can achieve low inequality and high environmental standards over time, or whether trade offs emerge in improving one dimension at the cost of another.

# Results

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4in]{fig_mfa_wb}
\caption{\textbf{Multiple factor analysis with World Bank data}  }
\label{fig:wb}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 4in]{fig_UN_ordination}
\caption{\textbf{Multiple factor analysis with United Nations data}  }
\label{fig:un}
\end{figure*}
```

Notes:

-   Paper figures:

    -   Plot with PCA selection (2-3 components?) and then change over time of the components.
        Do clustering on trayectories as well?
        -- and then add a map with the clusters for each dataset.

# References {.unnumbered}

::: {#refs}
:::

\pagebreak

# Supplementary Material {#sec:SM}

```{=tex}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{WB_environment_SDGs}
\caption{\textbf{World Bank environmental SDGs} }
\label{fig:SM_WB_env}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{WB_environment_SDGs_othervars}
\caption{\textbf{World Bank environmental SDGs additional variables} }
\label{fig:SM_WB_env2}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{WB_selected_vars}
\caption{\textbf{World Bank selected variables} }
\label{fig:SM_WB_select}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{WB_correlogram}
\caption{\textbf{Correlogram of World Bank selected variables} }
\label{fig:SM_WBcorr}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_SDG_10-14-15}
\caption{\textbf{United Nations SDGs 10, 14 and 25 dataset} }
\label{fig:SM_UNdata}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_selected_vars}
\caption{\textbf{UN selected variables} }
\label{fig:SM_UNselect}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_reduced_vars}
\caption{\textbf{UN reduced variables} }
\label{fig:SM_UNred}
\end{figure*}
```
```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{UN_correlogram}
\caption{\textbf{Correlogram of United Nations selected variables} }
\label{fig:SM_UNcorr}
\end{figure*}
```
